Automated Formal Synthesis of Provably Safe Digital Controllers for Continuous Plants

### 2019-07-20 automatica/paper.tex
all sections

A similar argument can be made for floating-point arithmetic.
We can thus disregard these steady-state errors (caused by
finite-precision arithmetic) when stability is ensured by synthesis, and then verify its
safety accounting for the finite-precision errors.

@@

We fix an index $k$, 
and we synthesize a safe controller by
unfolding the transition system (.e., the closed-loop model) $k$ steps and by selecting a controller $K$ and a single initial state, 
such that the states at each step do not violate the safety criteria (see Section~\ref{sec:safety}).    
That is, we ask the bounded model checker~\cite{ClarkeKL04} if there exists a $K$ that is safe for at least one $x_0$ in our set of all possible initial states, and a given fixed-point precision for the controller, and where the input signal remains within the specified bounds. 
This approach is sound, i.e., the controller produced is guaranteed to be safe, if the current $k$ is greater than the completeness threshold (see later step).  
We also assume some finite precision for the model and a given time discretisation, as described in section~\ref{sec:model}.  
We use a fixed-point precision in this description, given by $\langle I_p,F_p\rangle$, but if we are considering a floating-point controller we will instead model the plant with a floating-point precision, $\langle E_p,M_p\rangle$.
The checks that these assumptions hold are performed by the subsequent {\sc verify} stages.

@@

The first {\sc verify} stage, {\sc safety} is shown in Alg.~\ref{alg:safetycheck}. 
The algorithm checks that the candidate
solution $K$, which we have synthesized to be safe for at least one initial
state, is safe for \emph{all} possible initial states, i.e., it does not reach
the (complement) unsafe set within $k$ steps.  After unfolding the transition system corresponding
to the previously synthesized controller $k$ steps, we check that the safety
specification holds for any initial state.

@@

\begin{theorem}
  \label{th:safety}
\rev{Assuming that the initial set is included within the safe set, }
if a controller is safe for each of the corner cases of the hypercube of allowed initial states, 
i.e., the vertex states, then it is safe for any initial state in the hypercube. 
\end{theorem}
\begin{proof}
Consider the set of initial states, $X_0$, which we note is convex since it is a hypercube. 
Name $v_i$ its vertices, where $i=1,\ldots, 2^n$.  
Thus any point $x \in X_0$ can be expressed by convexity as $x = \sum_{i=1}^{2^n} \alpha_i v_i$, 
where $\sum_{i=1}^{2^n} \alpha_i =1$. Then if $x_0=x$, we obtain 
\begin{align*}
x_k   &= (A_d - B_d K)^k x = (A_d - B_d K)^k \sum_{i=1}^{2^n} \alpha_i v_i \\
      &= \sum_{i=1}^{2^n} \alpha_i (A_d - B_d K)^k v_i = \sum_{i=1}^{2^n} \alpha_i x_k^i, 
 \end{align*}     
%
where $x_k^i$ denotes the trajectories obtained from the single vertex
$v_i$.  We~conclude that any $k$-step trajectory is encompassed, within a
convex set, by those generated from the vertices. 
\rev{Recall that we have assumed that the initial set lies within the safe set, and that both are (convex) hyper-boxes. The conclusion follows. }
\qed
\end{proof}
Summarising, we only need to check $2^n$ initial states, 
where $n$ is the dimension of the state space (number of continuous variables).

@@

\begin{proof}
%
An asymptotically stable model is known to have converging dynamics.  
Assume that the eigenvalues of the closed-loop matrix are not repeated: this is sensible assumption to raise, 
since the eigenvalues are selected by the user. \rev{ The distance of the trajectory from the reference
point (which, for linear models, is the origin) decreases over time within subspaces related to real-valued
eigenvalues (say, $\theta < 0$): this can be shown considering the exponential $e^{\theta t} x_0$ ($x_0$ being the initial condition), 
which envelops the model dynamics within subspaces and is monotonically decreasing. } 
\rev{ However, this monotonic decrease cannot be ensured in general when dealing with complex eigenvalues. } 
Consider the closed-loop matrix that updates the
states in every discrete time step, and select the eigenvalue $\vartheta$ 
with the smallest (non-zero) imaginary part.  
Between any pair of consecutive time steps $k\,T_s$ and $(k+1)\,T_s$, the dynamics projected on
the corresponding eigenspace rotate of $\vartheta T_s$ radians.  Thus, taking
$\overline{k}$ as the ceiling of $\frac{2\pi}{\vartheta T_s}$, after
$k{\geq}\overline{k}$ steps the model trajectory has completed a full rotation \rev{within the relevant eigenspace}: 
this results in a point closer to the origin, \rev{as shown in Fig. \ref{fig:ct}}.  The synthesized $\overline{k}$ is thus a completeness threshold: 
\rev{indeed, since we have selected $\vartheta$ to be the complex eigenvalue with smallest imaginary part, 
any other pair of complex eigenvalues will correspond to dynamics that rotate faster within the corresponding eigenspace. 
Hence, $\overline{k}$ will again represent a completeness threshold. }
\qed 
%
\end{proof}

@@

In the second stage, 
the automated synthesis process starts with Step~A, where the
\tool~translates the model specification into an ANSI-C program.  
In Step~B, the discussed CEGIS engine is invoked, 
in order to synthesize the digital controller w.r.t.~the specification given on the model.  
Finally, in Step~C, the synthesized digital controller is generated.  
The output of \tool~is the closed-loop model with the synthesized digital controller, 
which is represented either as a transfer function or in state-space form (in this article we consider the latter case).  
The synthesis is considered to be \emph{successful} if a digital controller is
correctly synthesized with respect to the effects of finite-precision arithmetic within the 5 hour time-out.

Our CEGIS engine is implemented as an integrated module within the C bounded
model checker (CBMC)~\cite{ClarkeKL04}.  CBMC transforms the ANSI-C representation
of the closed-loop control model model into its internal representation (IR).  We
instrument this IR for each synthesis or verification scenario accordingly,
and use CBMC as an oracle to answer our queries.  CBMC itself relies on an
underlying SAT/SMT solver to address verification checks.  We~model the 
effects of finite-precision arithmetic 
explicitly using CBMC's nondeterminism API (e.g., \texttt{nondet},
\texttt{CPROVER\_assume} intrinsic functions).  Our module is included into
CBMC~5.7 and is available for 
download.\footnote{\url{https://github.com/diffblue/cbmc/archive/cbmc-5.7.zip}}
We have further created a VirtualBox OVA image with our tool pre-installed,
including our MATLAB-generated benchmarks and benchmark shell script, 
with instructions to run all benchmarks and to reproduce our
experiments.\footnote{\url{http://www.cprover.org/DSSynth/}}
%
%\textcolor{red}{This paragraph needs to be updated by Elizabeth and Dario.}
%
%The entire implementation consists of $XXX$ and $XXX$ lines of C code in the multi-stage and abstract
%accelerator back-ends, respectively, $XXXX$ lines of C code for utility functions such as interval, 
%fixed- and floating-point arithmetic, as well as $1000$ lines of C++ code in the control synthesis
%module in CEGIS CBMC.  We evaluate \tool~using $XXX$ SISO control system
%benchmarks.

@@

Our benchmark suite consists of 19 case studies extracted from the control literature~\cite{acrobot,cstr,CHEN1979389,KOKOTOVIC198023,gajic2008optimal,Franklin15, maglev, converters, gajic2008optimal}. These case studies comprise control models, which we represent in state space form as in \eqref{eq:ode1}. 
The models are time discretized, with sampling times ranging from 1s to 0.0001s. 
CEGIS initially starts with the coarsest discretisation (i.e., 1s sampling time), and if it fails to find a controller, reduces the sampling time. The initial states are bounded between $0.5$ and $-0.5$ for all states, and the safety specification requires that the states remain between $-1.5$ and $1.5$ \rev{(as remarked above, the initial set lies within the safe set and contains the origin)}. The input bounds are selected individually for each benchmark.
%\note{we need to at least qualitatively mention input selection, initial conditions, and also safe sets. how much can you say? }

@@

where the control is used to maintain the speed of an automobile constant, 
tracking a desired speed reference, and compensating disturbances and uncertainties.  % related to relief and terrain. 

@@

\textit{cstr}  and \textit{cstrtmp}  describe the pH dynamics of a reaction 

@@

describes a model for the cart position and the angle of an inverted pendulum placed on the cart,  

@@

which moves over a track through a DC motor.  

@@

The \textit{magsuspension} describes  the dynamic of a magnetic car suspension system. 

@@

orientation of a satellite with respect to an 
inertial frame during all the excursion. 

@@

Using the state-space models described in Section~\ref{exp:benchmarks}, 
the evaluation study has the following experimental goal:  
\begin{enumerate}
\item[] 
%[EG1] \textbf{(Performance)} 
Show that the multi-staged 
%and the abstraction-based 
CEGIS approach is able to generate finite-precision digital controllers using fixed-point and floating-point arithmetic in a reasonable amount of time.
%\item[EG2] \textbf{(Sanity check)} Confirm the stability \note{also?} and safety of the synthesized controllers outside 
%of our model representation using MATLAB. 
%\note{Where is this done?}
\end{enumerate}

@@

The mean run-time for the successful benchmarks is 225\,s for fixed-point and 125\,s for floating-point implementations, respectively.  
The median run-time for the successful benchmarks is 15.7\,s for fixed-point and 17.8\,s for floating-point implementations, respectively.
We consider these times to be short enough to be of practical use to control engineers, and thus affirm the validity of 
%EG1. 
the objective we have raised.

@@

%\textcolor{blue}{The synthesized controllers are confirmed to be safe outside of our model
%representation using MATLAB, achieving EG2 \note{please explain how this check is performed - is the check itself sound?}.  
A~link to the full experimental environment, including scripts to reproduce the results, all benchmarks and the tool, is provided in the
footnote as an Open Virtual Appliance (OVA).\footnote{\url{www.cprover.org/DSSynth}} 
The provided experimental environment runs multiple discretisations for each benchmark, and lists the fastest as the result synthesis time.

@@

\item[]
\textit{Plant model precision and discretization heuristics:} Our algorithm to
select suitable finite-precision arithmetic to model the plant behavior 
increases the precision by 8 bits at each step,  
in order to be compliant with the CBMC type API. 
Similarly, for time-discretization, we try a set of pre-defined time discretisations.
This works sufficiently well for our benchmarks, but
performance may suffer in some cases, for example if the completeness
threshold is high.

@@

We have presented an automated approach to synthesize provably-correct digital state-feedback controllers that ensure both stability and safety over state-space models.   
%The first approach relies on unfolding of the closed-loop model dynamics up to a completeness threshold, while the second
%one applies abstraction-refinement and acceleration to increase speed, whilst retaining soundness.
%
Our approach is novel within the digital control literature: 
we provide a fully automated synthesis method that is algorithmically and numerically sound,
considering various error sources in the implementation of the digital control algorithm and in the computational modeling of plant dynamics.
%
Our experimental results show that we are able to synthesize automatically safe controllers for diverse benchmarks from the digital control literature, 
within a reasonable amount of time.  
%
%\note{Let us collect here the main extensions we regard to be relevant: \\ 
%models with partial observations; \\ 
%implementation with SMT; \\ 
%continuous-time; \\ 
%more complex specs; \\ 
%\ldots }

>>>>

A similar argument can be made for floating-point arithmetic.
In conclusion, we can thus disregard these steady-state errors (caused by
finite-precision arithmetic) when stability is ensured by synthesis, and then verify  
safety accounting for the finite-precision errors.

@@

We fix an index $k$, 
and we synthesize a safe controller by
unfolding the transition system (i.e., the closed-loop model) $k$ steps and by selecting a controller $K$ and a single initial state, 
such that the states at each step do not violate the safety requirement (see Section~\ref{sec:safety}).    
That is, we ask the bounded model checker~\cite{ClarkeKL04} if there exists a $K$ that is safe for at least one $x_0$ in our set of all possible initial states, and a given fixed-point precision for the controller, and where the input signal remains within the specified bounds. 
This approach is sound, i.e., the controller produced is guaranteed to be safe, if the current $k$ is greater than the completeness threshold (see later step).  
We also assume some finite precision for the model and a given time discretisation, as described in section~\ref{sec:model}.  
We use a fixed-point precision in this description, given by $\langle I_p,F_p\rangle$, but if we are considering a floating-point controller we will instead model the plant with a floating-point precision, $\langle E_p,M_p\rangle$.
The checks that these assumptions hold are performed by the subsequent {\sc verify} stages.

@@

The first {\sc verify} stage, {\sc safety} is shown in Algorithm~\ref{alg:safetycheck}. 
The algorithm checks that the candidate
solution $K$, which we have synthesized to be safe for at least one initial
state, is safe for \emph{all} possible initial states, i.e., it does not reach
the (complement) unsafe set within $k$ steps.  After unfolding the transition system corresponding
to the previously synthesized controller $k$ steps, we check that the safety
specification holds for any initial state.

@@

\begin{theorem}
  \label{th:safety}
\rev{Assuming that the initial set is included within the safe set, }
if a controller is safe for each of the corner cases of the hyper-box of allowed initial states, 
i.e., the vertex states, then it is safe for any initial state in the hyper-box. 
\end{theorem}
\begin{proof}
Consider the set of initial states, $X_0$, which we note is convex since it is a hyper-box. 
Name $v_i$ its vertices, where $i=1,\ldots, 2^n$.  
Thus any point $x \in X_0$ can be expressed by convexity as $x = \sum_{i=1}^{2^n} \alpha_i v_i$, 
where $\sum_{i=1}^{2^n} \alpha_i =1$. Then if $x_0=x$, we obtain 
\begin{align*}
x_k   &= (A_d - B_d K)^k x = (A_d - B_d K)^k \sum_{i=1}^{2^n} \alpha_i v_i \\
      &= \sum_{i=1}^{2^n} \alpha_i (A_d - B_d K)^k v_i = \sum_{i=1}^{2^n} \alpha_i x_k^i, 
 \end{align*}     
%
where $x_k^i$ denotes the trajectories obtained from the single vertex
$v_i$.  We~conclude that any $k$-step trajectory is encompassed, within a
convex set, by those generated from the vertices. 
\rev{Recall that we have assumed that the initial set lies within the safe set, and that both are (convex) hyper-boxes. The conclusion follows. }
\qed
\end{proof}
Summarising, we only need to check $2^n$ initial states, 
where $n$ is the dimension of the state space (number of continuous variables). \todo{Add remark from reviewer that this number can be lower in actuality?}

@@

\begin{proof}
%
An asymptotically stable model is known to have converging dynamics.  
Assume that the eigenvalues of the closed-loop matrix are not repeated: this is sensible assumption to raise, 
since the eigenvalues are selected by the user. \rev{ The distance of the trajectory from the reference
point (which, for linear models, is the origin) decreases over time within subspaces related to real-valued
eigenvalues (say, $\theta < 0$): this can be shown considering the exponential $e^{\theta t} x_0$ ($x_0$ being the initial condition), 
which envelops the model dynamics within subspaces and is monotonically decreasing. } 
\rev{ However, this monotonic decrease cannot be ensured in general when dealing with complex eigenvalues. } 
In this second instance, consider the closed-loop matrix that updates the
states at each discrete time step, and select the eigenvalue $\vartheta$ 
with the smallest (non-zero) imaginary part.  
Between any pair of consecutive time steps $k\,T_s$ and $(k+1)\,T_s$, the dynamics projected on
the corresponding eigenspace rotate of $\vartheta T_s$ radians.  Thus, taking
$\overline{k}$ as the ceiling of $\frac{2\pi}{\vartheta T_s}$, after
$k{\geq}\overline{k}$ steps the model trajectory has completed a full rotation \rev{within the relevant eigenspace}: 
this results in a point closer to the origin, \rev{as shown in Fig. \ref{fig:ct}}.  The synthesized $\overline{k}$ is thus a completeness threshold: 
\rev{indeed, since we have selected $\vartheta$ to be the complex eigenvalue with smallest imaginary part, 
any other pair of complex eigenvalues will correspond to dynamics that rotate faster within the corresponding eigenspace. 
Hence, $\overline{k}$ will again represent a global completeness threshold. }
\qed 
%
\end{proof}

@@

In the second stage, 
the automated synthesis process starts with Step~A, where the
\tool~translates the model specification into an ANSI-C program.  
In Step~B, the discussed CEGIS engine is invoked, 
in order to synthesize the digital controller w.r.t.~the specification given on the model.  
Finally, in Step~C, the synthesized digital controller is generated.  
The output of \tool~is the closed-loop model with the synthesized digital controller, 
which is represented either as a transfer function or in state-space form (in this article we consider the latter case).  
The synthesis is considered to be \emph{successful} if a digital controller is
correctly synthesized with respect to the effects of finite-precision arithmetic within a time-out set to 5 hours.

The CEGIS engine is implemented as an integrated module within the C bounded
model checker (CBMC)~\cite{ClarkeKL04}.  CBMC transforms the ANSI-C representation
of the closed-loop control model model into its internal representation (IR).  We
instrument this IR for each synthesis or verification scenario accordingly,
and use CBMC as an oracle to answer our queries.  CBMC itself relies on an
underlying SAT/SMT solver to address verification checks.  We~model the 
effects of finite-precision arithmetic 
explicitly using CBMC's nondeterminism API (e.g., \texttt{nondet},
\texttt{CPROVER\_assume} intrinsic functions).  Our module is included into
CBMC~5.7 and is available for 
download.\footnote{\url{https://github.com/diffblue/cbmc/archive/cbmc-5.7.zip}}
We have further created a VirtualBox OVA image with our tool pre-installed,
including our MATLAB-generated benchmarks and benchmark shell script, 
with instructions to run all benchmarks and to reproduce our
experiments.\footnote{\url{http://www.cprover.org/DSSynth/}}
%
%\textcolor{red}{This paragraph needs to be updated by Elizabeth and Dario.}
%
%The entire implementation consists of $XXX$ and $XXX$ lines of C code in the multi-stage and abstract
%accelerator back-ends, respectively, $XXXX$ lines of C code for utility functions such as interval, 
%fixed- and floating-point arithmetic, as well as $1000$ lines of C++ code in the control synthesis
%module in CEGIS CBMC.  We evaluate \tool~using $XXX$ SISO control system
%benchmarks.

@@

Our benchmark suite consists of 19 case studies extracted from the control literature~\cite{acrobot,cstr,CHEN1979389,KOKOTOVIC198023,gajic2008optimal,Franklin15, maglev, converters, gajic2008optimal}. These case studies comprise control models, which we represent in state space form as in Equation \eqref{eq:ode1}. 
The models are time discretized, with sampling times ranging from 1s to 0.0001s. 
CEGIS initially starts with the coarsest discretisation (i.e., 1s sampling time), and if it fails to find a controller, it reduces the sampling time. 
The initial states are bounded between $0.5$ and $-0.5$ for all states, and the safety specification requires that the states remain between $-1.5$ and $1.5$ \rev{(as remarked above, the initial set lies within the safe set and contains the origin)}. The input bounds are selected individually for each benchmark.
%\note{we need to at least qualitatively mention input selection, initial conditions, and also safe sets. how much can you say? }

@@

where the control is used to maintain the speed of an automobile at a constant value (after tracking a desired speed reference), 
and compensating disturbances and uncertainties.  % related to relief and terrain. 

@@

The models \textit{cstr}  and \textit{cstrtmp}  describe the pH dynamics of a reaction 

@@

describes a model for the cart position and for the angle of an inverted pendulum placed on the cart,  

@@

which moves over a track by means of a DC motor.  

@@

The \textit{magsuspension} describes  the dynamics of a magnetic car suspension system. 

@@

orientation of a satellite with respect to an inertial frame. % during all the excursion. 

@@

Using the state-space models described in Section~\ref{exp:benchmarks}, 
the evaluation study has the following overall experimental goal:  
\begin{enumerate}
\item[] 
%[EG1] \textbf{(Performance)} 
Show that the multi-staged 
%and the abstraction-based 
CEGIS approach is able to generate finite-precision digital controllers using fixed-point and floating-point arithmetic in a reasonable amount of time.
%\item[EG2] \textbf{(Sanity check)} Confirm the stability \note{also?} and safety of the synthesized controllers outside 
%of our model representation using MATLAB. 
%\note{Where is this done?}
\end{enumerate}

@@

The mean run-time for the successful benchmarks is 225\,s for fixed-point and 125\,s for floating-point implementations, respectively.  
The median run-time for the successful benchmarks is 15.7\,s for fixed-point and 17.8\,s for floating-point implementations, respectively.
We consider these times to be short enough to be of practical use to control engineers, and thus assert the success of 
%EG1. 
the overall objective of the study that we have raised above.

%\textcolor{blue}{The synthesized controllers are confirmed to be safe outside of our model
%representation using MATLAB, achieving EG2 \note{please explain how this check is performed - is the check itself sound?}.  
A~link to the full experimental environment, including scripts to reproduce the results, all benchmarks and the tool, is provided in the
footnote as an Open Virtual Appliance (OVA).\footnote{\url{www.cprover.org/DSSynth}} 
The provided experimental environment runs multiple discretisations for each benchmark, and lists the fastest as the resulting synthesis time.

@@

@@

\item[]
\item[]
\textit{Plant model precision and discretization heuristics:} Our algorithm to
select suitable finite-precision arithmetic to model the plant behavior 
increases the precision by 8 bits at each step,  
in order to be compliant with the CBMC type API. 
Similarly, for time-discretization, we try a set of pre-defined time discretisations.
This works sufficiently well for our benchmarks, but
performance may suffer in some cases, for example if the completeness
threshold is high.

@@

We have presented an automated approach to synthesize provably-correct digital state-feedback controllers that ensure 
%both stability and 
safety over state-space models.   
%The first approach relies on unfolding of the closed-loop model dynamics up to a completeness threshold, while the second
%one applies abstraction-refinement and acceleration to increase speed, whilst retaining soundness.
%
Our approach is novel within the digital control literature: 
we provide a fully automated synthesis method that is algorithmically and numerically sound,
considering various error sources in the implementation of the digital control algorithm and in the computational modeling of plant dynamics.
%
Our experimental results show that we are able to synthesize automatically safe controllers for diverse benchmarks from the digital control literature, 
within a reasonable amount of time.  
%
%\note{Let us collect here the main extensions we regard to be relevant: \\ 
%models with partial observations; \\ 
%implementation with SMT; \\ 
%continuous-time; \\ 
%more complex specs; \\ 
%\ldots }
